My Findings

### 2024-06-10 Findings (Shared Folder)/Chapters/randomNotes.tex
new notes added

\begin{summary}
	In a nutshell, If we define 
	\[  J = \rho v \]
	where $ \rho :\Omega \times \R \to \R $ is a function representing the density, and $ v: \Omega \times \R \to \R^3 $ is the tangent vector to the orbit of particles generated by the action of group of diffeomorphisms that preserve the volume element $ \set{g^t}_{t\in T} $, then, for any surface $ \sigma \subset \Omega $, the rate of particles passing through the membrane will be
	\[ \frac{dN}{dt} = \int_\sigma J\cdot \hat{n}\ dA. \]
	Then by assuming that no particles will by destroyed to the thin air, and also no particles will born out of blue sky, then we can derive the continuity equation. In any volume $ V $ of our domain enclosed by the surface $ \sigma $, we can write
	\[ \frac{d}{dt} \int_V \rho\ dV = - \int_\sigma J\cdot \hat{n}\ dA = - \int_V \nabla\cdot J dV. \]
	By dropping the integration signs we will get
	\[ \frac{d\rho}{dt} + \nabla\cdot J = 0.  \]
\end{summary}

>>>>

\begin{summary}
	\label{summary:groupActionOfDiffeomorphisms}
	In a nutshell, If we define 
	\[  J = \rho v \]
	where $ \rho :\Omega \times \R \to \R $ is a function representing the density, and $ v: \Omega \times \R \to \R^3 $ is the tangent vector to the orbit of particles generated by the action of group of diffeomorphisms that preserve the volume element $ \set{g^t}_{t\in T} $, then, for any surface $ \sigma \subset \Omega $, the rate of particles passing through the membrane will be
	\[ \frac{dN}{dt} = \int_\sigma J\cdot \hat{n}\ dA. \]
	Then by assuming that no particles will by destroyed to the thin air, and also no particles will born out of blue sky, then we can derive the continuity equation. In any volume $ V $ of our domain enclosed by the surface $ \sigma $, we can write
	\[ \frac{d}{dt} \int_V \rho\ dV = - \int_\sigma J\cdot \hat{n}\ dA = - \int_V \nabla\cdot J dV. \]
	By dropping the integration signs we will get
	\[ \frac{d\rho}{dt} + \nabla\cdot J = 0.  \]
\end{summary}

\begin{observation}
	Reading the summary box above, one might conclude that the flux for diffusion might be something like $ J = - \rho \nabla\rho $ as the particles are moving opposite to their gradient. But this is wrong! In the case of diffusion, the velocity field is
	\[ v = -\frac{1}{\rho} \nabla \rho \]
	so the flux (Fick's law) will be of the form
	\[  J =  -\nabla\rho. \]
	At first, it was as very bizarre observation for me. I still do not understand how I can derive this velocity profile (i.e. flow velocity or simply flow).
\end{observation}

\section{Some cool PDE simulations}
Living in this level of abstraction, we can now do some stuff with PDEs that might look magic at the first glace. Here, I will show case some of these simulations.

\subsection{Diffusion and Advection}

\begin{quote}
	{\color{red} Be aware of the context.} Note that in this section, when we talk about the vector field $ F $, we are actually talking about the vector field that determines the advection term of the flux. I.e. we assume $ F $ is
	\[ J_\text{advection} = \rho F\]
\end{quote}
To have diffusion and advection at the same time, we need to design the flow in a way that can produce this. The first term of will be for the diffusion, where according to the observation box above will be of the form
\[  J = - \nabla\rho + \underbrace{\cdots}_{\text{advection term}} \]
To design the advection term, we need to observe that the particles should move along side some vector field. So it will be of the form
\[ \boxed{J_\text{advection} = \rho F(X)}  \]
where $ F: \Omega\to \R^3 $ is the desired vector field. So the advection-diffusion PDE will be 
\[ \frac{\partial \rho}{\partial t} = \nabla\cdot (-\nabla\rho + \rho F). \]
Now we can come up with all sort of creative vector fields.
\subsubsection{Simple velocity field}
For instance, for the vector field $ F(x,y)=(-1,0) $, we will have
\[ \frac{\partial \rho}{\partial t} = \nabla^2 \rho - \frac{\partial \rho}{\partial x}.  \]
Then the solution of this model will be like the following figure.
\begin{figure}[h!]
	\centering
	\includegraphics[width=0.7\linewidth]{images/simpleAdvectionDiffusion1}
	\caption{Advection diffusion with the vector field $ F(x,y,z) = (-1,0,0). $ The designed PDE forces the particles to move in this velocity field (or velocity profile).}
	\label{fig:simpleadvectiondiffusion1}
\end{figure}

\subsubsection{Particles Moving Down the Hill}
Here, we design a vector field that resembles particles moving down a potential hill to get to their minimum potential state. To design such a vector field, we simply first define an appropriate potential function and define the vector field to be its gradient. The potential that we will using for this purpose is 
\[ f(x,y) = 1/2(x^2 + y^2). \]
Then the corresponding vector field will be
\[ F(x,y) = -\nabla f(x,y) = (x,y). \]
So the desired PDE will be
\[ \frac{\partial \rho}{\partial t} = \nabla^2 \rho + x \frac{\partial \rho}{\partial x} + y \frac{\partial \rho}{\partial y} + \rho. \] 
Simulating this PDE we will get
\begin{figure}[h!]
	\centering
	\includegraphics[width=0.7\linewidth]{images/movingToCenter.png}
	\caption{Advection diffusion with the vector field $ F(x,y) = (x,y)$ which is the gradient of potential function $ f(x,y) = 1/2(x^2 + y^2) $.}
	\label{fig:potentialAsGradient}
\end{figure}

\subsubsection{Moving Around the Origin}
What if at each point, we rotate the vector vector of the previous example by 90 degrees so that the particles will move around the origin. The vector field in previous example was $ F(x,y) = (x,y) $. To do the rotation we need ot multiply that in the rotation matrix
\[ F_\text{new}(x,y)=  \matt{0}{-1}{1}{0}F(x,y) = \vectt{-y}{x}.  \]
Thus the desired PDE will be
\[ \frac{\partial \rho}{\partial t} = \nabla^2 \rho - y\frac{\partial \rho}{\partial x} + x \frac{\partial \rho}{\partial y}. \]
By simulating this PDE we will get
\begin{figure}[h!]
	\centering
	\includegraphics[width=0.7\linewidth]{images/movingAround.png}
	\caption{Advection diffusion with the vector field $ F(x,y,z) = (-y,x)$. This vector field is derived by rotating the vectors in the previous example by 90 degrees. Surprisingly, there are no potentials that $ F $ ce be its gradient! In other words, we can not derive $ F $ from any potential.}
	\label{fig:simpleadvectiondiffusion1}
\end{figure}

\subsubsection{Spiraling in to the Origin}
We can start with the vector field $ F(x,y) = (x,y) $ and this time, rather time rotating it with a rotation matrix, we can rotate that using a matrix corresponding to a complex number $ z = \mu + i \omega $, i.e.
\[ F_\text{new}(x,y) = \matt{\mu}{-\omega}{\omega}{\mu}F(x,y). \]
We can take this complex number to be $ z = -1 + i $. Thus our new vector field will be
\[ F_\text{new}(x,y) = (-x+y, -x-y).  \]
So our PDE will read
\[ \frac{\partial \rho}{\partial t} = \nabla^2 \rho + (-x+y)\frac{\partial \rho}{\partial x} + (-x-y)\frac{\partial \rho}{\partial y} - 2\rho. \]
Simulating this PDE we will get

\begin{figure}[h!]
	\centering
	\includegraphics[width=0.5\linewidth]{images/spiralAdvectionDiffusion.png}
	\caption{Advection diffusion with the vector field $ F_{\text{advection}}(x,y,z) = (-x+y, -x-y)$.}
	\label{fig:spiralAdvectionDiffusion}
\end{figure}
\FloatBarrier

Now the interesting question is why do we see this? To observe this, remember that the velocity profile is the same as the vector field that we specify for the model (in this case $ F_\text{new}(x,y) $). On the other hand, we know that the velocity profile is in fact the right hand side of the ODE for flow
\[ \frac{d \phi}{d t} = F_\text{new}(x,y). \]
Note that $ \phi $ is flow, or following the terminology of \autoref{summary:groupActionOfDiffeomorphisms}, it is one element of the group diffeomorphisms that preserve the volume element (in the that terminology we denote it by $ g^t $). For the simulation above, ODE for $ \phi:\R^2 \to \R^2 $ is 
\[ \vectt{\dot{\phi_1}}{\dot{\phi_2}} = \matt{-1}{1}{-1}{-1}\vectt{x}{y}. \]
The phase plane of flow will be

\begin{figure}[h!]
	\centering
	\includegraphics[width=0.2\linewidth]{images/flowLinePlot.png}
	\caption{The phase plane of flow $ \phi_\text{advection} = (\phi_1,\phi_2) $}
	\label{fig:SpiralAdvectionDiffusionFlowPhasePortrait}
\end{figure}
\FloatBarrier

The following summary box summarizes our observation.

\begin{summary}[Advection-Diffusion PDE]
	In an advection diffusion PDE for $ \rho $ the flux will have two terms, 
	\[ J = \underbrace{-D\nabla \rho}_{J_\text{diffusion} = \rho v_\text{diffusion}}+ \underbrace{\rho F}_{J_\text{advection} = \rho v_\text{advection}} \]
	where $ F $ is a vector field, which is the velocity profile of advection (i.e. $ v_\text{advection} = F $). Thus the flow (or in the terminology of \autoref{summary:groupActionOfDiffeomorphisms} the diffeomorphism preserving the volume element) will be given with the following ODE
	\[ \frac{d \phi_\text{advection}}{dt} = v_\text{advection}. \]
	In the cases where we have small diffusion, the solution of the advection-diffusion PDE will look like the phase plane diagram of $ \phi_\text{advection} $. Compare \autoref{fig:spiralAdvectionDiffusion} with  \autoref{fig:SpiralAdvectionDiffusionFlowPhasePortrait} to see this.
\end{summary}

\begin{observation}
	Given what we discussed above, I have the following idea to study the phase plane of an ODE system in a wise way! Consider the following ODE system
	\[ \dot{\Phi} = F(\Phi). \]
	In the corresponding phase space diagram, we usually plot the vector field $ F(X) $ and roughly draw some important orbits (see \autoref{fig:SpiralAdvectionDiffusionFlowPhasePortrait}). However, using the machinery we discussed in the summary box above, we can also evaluate what will happen for a given volume of initial positions (i.e. a volume of phase space). We can design a PDE whose flow is given by $ \Phi $. Thus this PDE will be
	\[ \frac{\partial \rho}{\partial t} = -\nabla\cdot(\rho F(\Phi)), \]
	where $ \rho $ represents the density of states in the phase space. Note that the equation above is nothing more that just inserting an appropriate choice of flux into the continuity equation.
\end{observation}

\begin{observation}[A proof for Liouville's theorem (Hamiltonian)]
	After doing some research, I found out that the approach I have had in the observation above can be used to prove the Liouville's theorem for Hamiltonian. This was an independent discovery and proof of this theorem!

	\begin{proof}
		consider the following Hamiltonian system
		\[ \Phi = \vectt{\dot{q}}{\dot{p}} = \vectt{\partial H/ \partial p}{- \partial H / \partial q} = F(\Phi). \]
		Let $ \rho $ be the density of states in the phase plane, since all of these points are following the flow profile, thus the corresponding flux will be
		\[  J = \rho \Phi.  \]
		Inserting this into the continuity equation we will get
		\[ \frac{\partial \rho}{\partial t} = -\nabla\cdot (\rho \Phi) = - \nabla \rho \cdot \Phi - \rho\nabla\cdot\Phi . \]
		For the first term in the right hand side we have
		\[ \nabla\rho\cdot \phi = \frac{\partial \rho}{\partial q}\cdot{q} + \frac{\partial \rho}{\partial p} \cdot{p}. \]
		And for the second term we can write
		\[ \rho\nabla\cdot\Phi = \rho(\frac{\partial^2 H}{\partial q\partial p} - \frac{\partial^2 H}{\partial p \partial q}) = 0. \]
		Thus we can write
		\[ \frac{\partial\rho}{\partial t} + (\frac{\partial \rho}{\partial q}\cdot{q} + \frac{\partial \rho}{\partial p} \cdot{p}) = 0.  \]
		This completes the proof.
	\end{proof}
\end{observation}

\begin{observation}
	The time evolution of the density of a die in an \emph{incompressible fluid} due to \emph{advection}, i.e. with the flux given by
	\[  J=\rho v\]
	 is analogous to the time evolution of the density of states in the phase plane of its flow $ \Phi $ (determined by the equation $ \dot{\Phi}=v $).
\end{observation}

In the observation box above, we have only talked about the rationale behind the PDEs that contains only the advection term. But what about the diffusion term? What are the ODE systems where the density of states in its phase plane will have a diffusion like behaviour? The following important observation box answers this important question.

\begin{observation}[Diffusion of the density of states in phase space]
	The statement of the question is as follows.
	\begin{quote}
		I was studying the PDEs governing the time evolution of the concentration of some die in an incompressible fluid. I noticed this general framework behind that which can be applied to the time evolution of the density of states in the phase plane of an ODE system. Once we know the governing ODE for the state of a system, lets say
		\[ \dot{X}  = F(X) \] 
		Then we can draw the phase plane with some important orbits in it. However, in order to find out what is the time evolution of the density of states, then we can easily find a corresponding PDE given by first defining $  J = \rho F$ and then writing
		\[ \rho_t = -\nabla\cdot(\rho F) = -\nabla\cdot(J). \]
		This gives the time evolution of a given density of states. In order to have a diffusion like behaviour in the phase plane, we should have $  J = \rho F = -\nabla \rho$. This means that the corresponding RHS for ODE should be $ F = -1/\rho \nabla \rho $ which does not make sense for me, as when writing the ODE for $ X $ the right hand side depends on the density of the states which does not makes sense in an ODE. So my question is that in what situation the density of states in and ODE system will have a diffusion like behaviour?
	\end{quote}

	The answer to this question has very deep roots in the stochastic vs. deterministic systems. The fact is that in \emph{deterministic} ODEs, we will \emph{never} observe a \emph{diffusion like behaviour} in the density of states of phase plane. The diffusion like behaviour of the density of states happens only in the case of Stochastic Differential Equations (SDE). As in the case of deterministic ODE systems, defining a suitable flux and inserting it into the continuity equation gives us the PDE for the time evolution of the density of states, doing a similar process gives us the Fokker-Plank equation that describes the time evolution of the density of states (which is called probability density function in this context) of the corresponding SDE. Consider the following SDE
	\[ dX_t = \mu(X_t,t)dt + \sigma(X_t,t)dW_t, \]
	where $ \mu(X_t,t) $ is the drift and $ D(X_t,t) = \sigma^2(X_t,t)/2 $ is the diffusion coefficient. The corresponding Fokker-Plank equation for the probability density function $ p(x,t) $ of the random variable $ X_t $ (analogous to the notion of the density of states in what we had discussed above) is given by
	\[ \frac{\partial}{\partial t} p(x,t) = -\frac{\partial}{\partial x}[\mu(x,t)p(x,t) - \frac{\partial}{\partial x}(D(x,t)p)]. \]
	Similarly, for the higher dimensions, the SDE will be of the form
	\[ d\mathbf{X_t} = \boldsymbol{\mu}(\mathbf{X_t},t)dt + \boldsymbol{\sigma}(\mathbf{X_t},t)d\mathbf{W_t}, \]
	where $ \mathbf{X_t} $ and $ \boldsymbol{\mu} $ are $ N $ dimensional vectors, and $ \boldsymbol{\sigma} $ is $ N\times M $ matrix, and $ \mathbf{W_t} $ is an $ M $ dimensional Wiener process. Then the corresponding Fokker-Plank equation will be given as
	\[ \frac{\partial p}{\partial t} = -\nabla\cdot(p\boldsymbol{\mu} + \boldsymbol{\nabla}\cdot(p\mathbf{D})) \]
	where the bold letter $ \boldsymbol{\nabla}\cdot$  applied on a matrix outputs a vector where each row is the divergence of the rows of the input matrix, and the matrix $ \mathbf{D} = \frac{1}{2}\boldsymbol{\sigma}\boldsymbol{\sigma^T} $. See examples below for more details.
\end{observation}

\begin{example}[1D Diffusion Equation]
	Consider the following SDE
	\[ dX_t = \sqrt{D}dW_t. \]
	The PDE governing the density of states (or the probability density function), i.e. the Fokker-Plank equation is given by
	\[ \frac{\partial p}{\partial t} = -\frac{\partial}{\partial x} (-D \frac{\partial p}{\partial x}) = D\frac{\partial ^2 p}{\partial x^2}, \]
	which gives rise to the 1D diffusion equation.
\end{example}

\begin{example}[2D Diffusion Equation]
	Consider the following system of SDEs
	\[ \vectt{dX_t}{dY_t} = \matt{1}{0}{0}{1}\vectt{dW^{[1]}_t}{dW^{[2]}_t}. \]
	The corresponding Fokker-Plank equation will be
	\[ \frac{\partial p}{\partial x} = \nabla\cdot(\boldsymbol{\nabla}\cdot\matt{p}{0}{0}{p}) 
	= \nabla\cdot(\vectt{p_x}{p_y}) = p_{xx} + p_{yy},
	 \]
	Which is the 2D diffusion equation.
\end{example}

\begin{example}[Advection Diffusion Equation]
	For instance, if we have
	\[ \vectt{dX_t}{dY_t} = \mathbf{F}(X_t)dt +  \matt{1}{0}{0}{1}\vectt{dW^{[1]}_t}{dW^{[2]}_t}. \]
	Then the corresponding Fokker-Plank equation will be
	\[ \frac{\partial p}{\partial t} = -\nabla\cdot( -\boldsymbol{\nabla}\cdot(\matt{p}{0}{0}{p}) + p\mathbf{F}) = -\nabla\cdot(-\nabla p + p\mathbf{F}). \]
	which is nothing but the familiar advection-diffusion system. Note that first term inside the parenthesis is the Fick's first law and the second term is taking care of the fact that the particles are advecting according to the vector field $ \mathbf{F} $.
\end{example}

\begin{example}[Boltzmann Distribution at the Thermodynamic Equilibrium]
	Consider the following SDE that is the overdamped Langevin equation
	\[ dX_t = -\frac{1}{k_BT}(\nabla U) dt + dW_t. \]
	Observe that similar to the example corresponding to \autoref{fig:potentialAsGradient} we defined the vector field for advection (drift) term as the gradient of a potential function. Thus similar to the example mentioned above, the particles will following trajectories down the potential well. The corresponding Fokker-Plank equation will be
	\[ \frac{\partial p}{\partial t} = -\frac{\partial}{\partial x}(-\frac{\partial p}{\partial x} - p\frac{1}{k_BT}\nabla U ), \]
	where we have kept writing the derivative of $ U $ as $ \nabla U $ to emphasis it is the gradient of a potential, but in fact we have $ \nabla U = \frac{\partial U}{\partial x} $. The steady state solution for the PDE above is
	\[ \frac{d p}{d x} = \frac{-p}{k_BT}\frac{d U}{d x}. \]
	The solution of the ODE above is given by
	\[\boxed{ p(x) = A e^{-\frac{U(x)}{k_BT}}}. \]
\end{example}

\begin{example}[Ornstein-Uhlenbeck process]
	Consider the following SDE
	\[ dX_t = -X_tdt + dW_t. \]
	Then the corresponding Fokker-Plank equation will be
	\[ \frac{\partial p}{\partial t} = -\frac{\partial}{\partial x}(-px - p_x). \]
	The steady state solution of this PDE will be given as
	\[ \frac{d p}{dx}  = - p x + c.\]
	The solution of this differential equation is given as
	\[ \boxed{p(x) = A e^{-x^2/2}}. \]
	This model is used to model the financial markets.
\end{example}
